A Logical Solution of the Linearity Challenge

### 2025-12-02 paper/Congruence.lagda.tex
included reduction
restored congruence

\begin{code}[hide]
open import Data.Bool using (Bool; true; false; if_then_else_)
open import Data.Product using (_×_; _,_; ∃; ∃-syntax)
open import Relation.Binary.PropositionalEquality using (_≡_; refl; cong; cong₂)
open import Data.List.Base using (List; []; _∷_; [_]; _++_)

@@

\subsection{Operational Semantics}
\label{sec:semantics-agda}

@@

\begin{AgdaAlign}
\begin{code}
data _⊒_ : ∀{Γ} → Process Γ → Process Γ → Set where
\end{code}

@@

\begin{code}
  s-comm :
    ∀{Γ Γ₁ Γ₂ A B P Q} (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) →
    cut d p P Q ⊒ cut (dual-symm d) (+-comm p) Q P
\end{code}

@@

\begin{code}[hide]
  s-link :
    ∀{Γ A B}
    (d : Dual A B) (p : Γ ≃ [ A ] + [ B ]) →
    link d p ⊒ link (dual-symm d) (+-comm p)

  s-fail :
    ∀{Γ Γ₁ Γ₂ Δ A B P} (d : Dual A B)
    (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ ⊤ , Δ) →
    let _ , _ , q′ = +-assoc-l p q in
    cut d p (fail (split-r q)) P ⊒ fail q′
\end{code}

@@

\begin{code}
  s-wait :
    ∀{Γ Γ₁ Γ₂ Δ A B} {P : Process (A ∷ Δ)} {Q : Process (B ∷ Γ₂)}
    (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ ⊥ , Δ) →
    let _ , p′ , q′ = +-assoc-l p q in
    cut d p (wait (split-r q) P) Q ⊒ wait q′ (cut d p′ P Q)
\end{code}

@@

\begin{code}
  s-select-l :
    ∀{Γ Γ₁ Γ₂ Δ A B C D} {P : Process (C ∷ A ∷ Δ)} {Q : Process (B ∷ Γ₂)}
    (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ C ⊕ D , Δ) →
    let _ , p′ , q′ = +-assoc-l p q in
    cut d p (select true (split-r q) P) Q ⊒
    select true q′ (cut d (split-l p′) (#process #here P) Q)
\end{code}

@@

\begin{code}[hide]
  s-select-r :
    ∀{Γ Γ₁ Γ₂ Δ A B C D}
    {P : Process (D ∷ A ∷ Δ)}
    {Q : Process (B ∷ Γ₂)}
    (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ C ⊕ D , Δ) →
    let _ , p′ , q′ = +-assoc-l p q in
    cut d p (select false (split-r q) P) Q ⊒
    select false q′ (cut d (split-l p′) (#process #here P) Q)

  s-case :
    ∀{Γ A B A₁ A₂ Γ₁ Γ₂ Δ}
    {P : Process (A₁ ∷ A ∷ Δ)}
    {Q : Process (A₂ ∷ A ∷ Δ)}
    {R : Process (B ∷ Γ₂)}
    (d : Dual A B)
    (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ A₁ & A₂ , Δ) →
    let _ , p′ , q′ = +-assoc-l p q in
    cut d p (case (split-r q) P Q) R ⊒
    case q′ (cut d (split-l p′) (#process #here P) R)
            (cut d (split-l p′) (#process #here Q) R)
\end{code}

@@

\begin{code}
  s-fork-l :
    ∀{Γ Γ₁ Γ₂ Δ Δ₁ Δ₂ A B C D}
    {P : Process (C ∷ A ∷ Δ₁)} {Q : Process (D ∷ Δ₂)} {R : Process (B ∷ Γ₂)}
    (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ C ⊗ D , Δ) (r : Δ ≃ Δ₁ + Δ₂) →
    let _ , p′ , q′ = +-assoc-l p q in
    let _ , p′′ , r′ = +-assoc-l p′ r in
    let _ , q′′ , r′′ = +-assoc-r r′ (+-comm p′′) in
    cut d p (fork (split-r q) (split-l r) P Q) R ⊒
    fork q′ r′′ (cut d (split-l q′′) (#process #here P) R) Q
\end{code}

Recall from \Cref{tab:semantics} that we want to apply this rule on a process of
the form $\Cut\x{\Fork\y\u\v{P}{Q}}{R}$ when $x \in \fn{P}$. We capture this
condition by means of the splitting $(\AgdaInductiveConstructor{split-l}~r)$,
meaning that the type $A$ of $x$ ends up in the typing context for $P$ and not
in the one for $Q$.
%
The symmetric rule \SForkR is modeled by another constructor
\AgdaInductiveConstructor{s-fork-r}, which is not shown here but is similar to
\AgdaInductiveConstructor{s-fork-l} except that
$(\AgdaInductiveConstructor{split-l}~r)$ is replaced by
$(\AgdaInductiveConstructor{split-r}~r)$.

\begin{code}[hide]
  s-fork-r :
    ∀{Γ Γ₁ Γ₂ Δ Δ₁ Δ₂ A B C D}
    {P : Process (C ∷ Δ₁)}
    {Q : Process (D ∷ A ∷ Δ₂)}
    {R : Process (B ∷ Γ₂)}
    (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ C ⊗ D , Δ)
    (r : Δ ≃ Δ₁ + Δ₂) →
    let _ , p′ , q′ = +-assoc-l p q in
    let _ , p′′ , r′ = +-assoc-l p′ r in
    cut d p (fork (split-r q) (split-r r) P Q) R ⊒
    fork q′ r′ P (cut d (split-l p′′) (#process #here Q) R)

  s-join :
    ∀{Γ Γ₁ Γ₂ Δ A B C D}
    {P : Process (D ∷ C ∷ A ∷ Δ)}
    {Q : Process (B ∷ Γ₂)}
    (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ C ⅋ D , Δ) →
    let _ , p′ , q′ = +-assoc-l p q in
    cut d p (join (split-r q) P) Q ⊒
    join q′ (cut d (split-l (split-l p′)) (#process #rot P) Q)

  s-server :
    ∀{Γ A B C Γ₁ Γ₂ Δ₁}
    {P : Process (C ∷ ¿ A ∷ Δ₁)}
    {Q : Process (B ∷ Γ₂)}
    (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ ¡ C , Δ₁) (r : Γ₂ ≃ [] + Γ₂)
    (un₁ : Un Δ₁) (un₂ : Un Γ₂) →
    let _ , p′ , q′ = +-assoc-l p q in
    cut (d-?-! d) p (server (split-r q) (un-∷ un₁) P) (server (split-l r) un₂ Q) ⊒
    server q′ (#un+ p′ un₁ un₂) (cut (d-?-! d) (split-l p′) (#process #here P) (server (split-l r) un₂ Q))

  s-client :
    ∀{Γ A B C Γ₁ Γ₂ Δ}
    {P : Process (C ∷ A ∷ Δ)}
    {Q : Process (B ∷ Γ₂)}
    (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ ¿ C , Δ) →
    let _ , p′ , q′ = +-assoc-l p q in
    cut d p (client (split-r q) P) Q ⊒
    client q′ (cut d (split-l p′) (#process #here P) Q)

  s-weaken :
    ∀{Γ A B C Γ₁ Γ₂ Δ}
    {P : Process (A ∷ Δ)}
    {Q : Process (B ∷ Γ₂)}
    (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ ¿ C , Δ) →
    let _ , p′ , q′ = +-assoc-l p q in
    cut d p (weaken (split-r q) P) Q ⊒
    weaken q′ (cut d p′ P Q)

  s-contract :
    ∀{Γ A B C Γ₁ Γ₂ Δ}
    {P : Process (¿ C ∷ ¿ C ∷ A ∷ Δ)}
    {Q : Process (B ∷ Γ₂)}
    (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ ¿ C , Δ) →
    let _ , p′ , q′ = +-assoc-l p q in
    cut d p (contract (split-r q) P) Q ⊒
    contract q′ (cut d (split-l (split-l p′)) (#process #rot P) Q)
\end{code}

@@

\begin{code}
  s-refl  : ∀{Γ} {P : Process Γ} → P ⊒ P
  s-tran  : ∀{Γ} {P Q R : Process Γ} → P ⊒ Q → Q ⊒ R → P ⊒ R
  s-cong  : ∀{Γ Γ₁ Γ₂ A A′} {P Q : Process (A ∷ Γ₁)} {P′ Q′ : Process (A′ ∷ Γ₂)}
            (d : Dual A A′) (p : Γ ≃ Γ₁ + Γ₂) → P ⊒ Q → P′ ⊒ Q′ →
            cut d p P P′ ⊒ cut d p Q Q′
\end{code}

>>>>

\begin{code}[hide]
{-# OPTIONS --rewriting #-}
open import Data.Product using (_×_; _,_; ∃; ∃-syntax)
open import Data.List.Base using (List; []; _∷_; [_]; _++_)

@@

\subsection{Structural Pre-Congruence}
\label{sec:congruence-agda}

@@

\begin{AgdaAlign}
\begin{code}
data _⊒_ {Γ} : Process Γ → Process Γ → Set where
\end{code}

@@

\begin{code}
  s-comm :
    ∀{A Γ₁ Γ₂ P Q} (p : Γ ≃ Γ₁ + Γ₂) →
    cut {A} p P Q ⊒ cut (+-comm p) Q P
\end{code}

@@

implicit rewriting

\begin{code}[hide]
  s-link :
    ∀{A} (p : Γ ≃ [ A ] + [ dual A ]) → link p ⊒ link (+-comm p)
  s-fail :
    ∀{A Γ₁ Γ₂ Δ P} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ ⊤ ⊳ Δ) →
    let _ , _ , q′ = +-assoc-l p q in
    cut {A} p (fail (> q)) P ⊒ fail q′
\end{code}

@@

\begin{code}
  s-wait :
    ∀{Γ₁ Γ₂ Δ A P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ ⊥ ⊳ Δ) →
    let _ , p′ , q′ = +-assoc-l p q in
    cut {A} p (wait (> q) P) Q ⊒ wait q′ (cut p′ P Q)
\end{code}

@@

\begin{code}[hide]
  s-case :
    ∀{A B C Γ₁ Γ₂ Δ P Q R} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ B & C ⊳ Δ) →
    let _ , p′ , q′ = +-assoc-l p q in
    cut {A} p (case (> q) P Q) R ⊒ case q′ (cut (< p′) (↭process swap P) R)
                                           (cut (< p′) (↭process swap Q) R)
\end{code}
\begin{code}
  s-left :
    ∀{Γ₁ Γ₂ Δ A B C P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ B ⊕ C ⊳ Δ) →
    let _ , p′ , q′ = +-assoc-l p q in
    cut {A} p (left (> q) P) Q ⊒ left q′ (cut (< p′) (↭process swap P) Q)
\end{code}
\begin{code}[hide]
  s-right :
    ∀{Γ₁ Γ₂ Δ A B C P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ B ⊕ C ⊳ Δ) →
    let _ , p′ , q′ = +-assoc-l p q in
    cut {A} p (right (> q) P) Q ⊒ right q′ (cut (< p′) (↭process swap P) Q)
\end{code}

@@

@@

\begin{code}[hide]
  s-join :
    ∀{Γ₁ Γ₂ Δ A B C P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ B ⅋ C ⊳ Δ) →
    let _ , p′ , q′ = +-assoc-l p q in
    cut {A} p (join (> q) P) Q ⊒
    join q′ (cut (< < p′) (↭process (↭shift {A} {C ∷ B ∷ []}) P) Q)
\end{code}
\begin{code}
  s-fork-l :
    ∀{Γ₁ Γ₂ Δ Δ₁ Δ₂ A B C P Q R}
    (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ B ⊗ C ⊳ Δ) (r : Δ ≃ Δ₁ + Δ₂) →
    let _ , p′ , q′ = +-assoc-l p q in
    let _ , p′′ , r′ = +-assoc-l p′ r in
    let _ , q′′ , r′′ = +-assoc-r r′ (+-comm p′′) in
    cut {A} p (fork (> q) (< r) P Q) R ⊒
    fork q′ r′′ (cut (< q′′) (↭process swap P) R) Q
\end{code}

Recall from \Cref{tab:semantics} that we want to apply this rule on a process of
the form $\Cut[A]\x{\Fork\y\u\v{P}{Q}}{R}$ when $x \in \fn{P}$. We capture this
condition by means of the splitting $(\AgdaInductiveConstructor{<}~r)$, meaning
that the type $A$ of $x$ ends up in the typing context for $P$ and not in the
one for $Q$.
%
The symmetric rule \SForkR is modeled by another constructor
\AgdaInductiveConstructor{s-fork-r}, not shown here but similar to
\AgdaInductiveConstructor{s-fork-l} except that
$(\AgdaInductiveConstructor{<}~r)$ is replaced by
$(\AgdaInductiveConstructor{>}~r)$.

\begin{code}[hide]
  s-fork-r :
    ∀{Γ₁ Γ₂ Δ Δ₁ Δ₂ A B C P Q R}
    (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ B ⊗ C ⊳ Δ) (r : Δ ≃ Δ₁ + Δ₂) →
    let _ , p′ , q′ = +-assoc-l p q in
    let _ , p′′ , r′ = +-assoc-l p′ r in
    cut {A} p (fork (> q) (> r) P Q) R ⊒
    fork q′ r′ P (cut (< p′′) (↭process swap Q) R)
  s-all :
    ∀{A B Γ₁ Γ₂ Δ F Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ `∀ B ⊳ Δ) ->
    let _ , p' , q' = +-assoc-l p q in
    cut {A} p (all (> q) F) Q ⊒ all q' λ σ → cut (< p') (↭process swap (F σ)) Q
  s-ex :
    ∀{A B C Γ₁ Γ₂ Δ P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ `∃ B ⊳ Δ) ->
    let _ , p' , q' = +-assoc-l p q in
    cut {A} p (ex {_} {C} (> q) P) Q ⊒ ex q' (cut (< p') (↭process swap P) Q)
  s-server :
    ∀{A B Γ₁ Γ₂ Δ₁ P Q}
    (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ `! B ⊳ Δ₁) (r : Γ₂ ≃ [] + Γ₂)
    (un₁ : Un Δ₁) (un₂ : Un Γ₂) →
    let _ , p′ , q′ = +-assoc-l p q in
    cut {`? A} p (server (> q) (un-∷ un₁) P) (server (< r) un₂ Q) ⊒
    server q′ (+-un p′ un₁ un₂) (cut (< p′) (↭process swap P) (server (< r) un₂ Q))
  s-client :
    ∀{A B Γ₁ Γ₂ Δ P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ `? B ⊳ Δ) →
    let _ , p′ , q′ = +-assoc-l p q in
    cut {A} p (client (> q) P) Q ⊒ client q′ (cut (< p′) (↭process swap P) Q)
  s-weaken :
    ∀{A B Γ₁ Γ₂ Δ P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ `? B ⊳ Δ) →
    let _ , p′ , q′ = +-assoc-l p q in
    cut {A} p (weaken (> q) P) Q ⊒ weaken q′ (cut p′ P Q)
  s-contract :
    ∀{A B Γ₁ Γ₂ Δ P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ `? B ⊳ Δ) →
    let _ , p′ , q′ = +-assoc-l p q in
    cut {A} p (contract (> q) P) Q ⊒
    contract q′ (cut (< < p′) (↭process (↭shift {A} {`? B ∷ `? B ∷ []}) P) Q)
\end{code}

@@

\begin{code}
  s-refl  : ∀{P} → P ⊒ P
  s-tran  : ∀{P Q R} → P ⊒ Q → Q ⊒ R → P ⊒ R
  s-cong  : ∀{Γ₁ Γ₂ A P Q P′ Q′} (p : Γ ≃ Γ₁ + Γ₂) →
            P ⊒ Q → P′ ⊒ Q′ → cut {A} p P P′ ⊒ cut p Q Q′
\end{code}
